Well-quasi-orderings on word languages

### 2024-08-09 src/infixes.tex
proving the regular language case

    Conversely, let us prove that \cref{infix-embedding-count:item} implies
    \cref{infix-embedding-embeds:item}. To that end, let us first define
    inductively a sequence $\seqof{X_n}$ of subsets of $X$ as follows: $X_0$ is
    the set of minimal elements of $X$, and $X_{n+1}$ is the set of minimal
    elements of $X \setminus \bigcup_{i \leq n} X_i$. Because initial segments
    of $X$ are finite, it is clear that $X = \bigcup_{n \in \Nat} X_n$. To
    simplify notations, we will also use $Y_n \defined \bigcup_{i \leq n} X_i$.
    We also let $L_0 \defined a b^* a$, and 
    $L_{n+1} \defined L_n \cup c^{n+1} a^{n+1} b^* a^{n+1} c^{n+1} (L_n c^{n+1})^*$.
    Now, let us inductively define \kl{embeddings} $f_n \colon Y_n \to
    L_n$.

    For the base case, let $\seqof{x_i}[i \in \Nat]$ be an enumeration of
    $X_0$, which is possible because $X$ is countable. We define $f_0 \colon
    X_0 \to \Sigma^*$ as follows: $f_0(x_i) \defined a b^i a$. It is immediate
    that $f_0$ is an \kl{embedding}. For the inductive step, let $\seqof{x_i}[i
    \in \Nat]$ be an enumeration of $X_{n+1}$, and assume that $f_n$ is an
    \kl{embedding} from $Y_n$ to $\Sigma^*$.
    Let us define $D_i \defined \setof{ y \in Y_n }{ y \preceq x_i }$
    for $i \in \Nat$, all of which are finite by assumption.
    We then let for all $i \in \Nat$:
    \begin{equation*}
        f_{n+1}(x_i) \defined 
        \underbrace{c^{n+1} a^{n+1} b^i a^{n+1} c^{n+1} \prod_{y \in D_i} \overbrace{f_n(y)}^{\in L_n} c^{n+1}}_{\in L_{n+1}}
        \quad .
    \end{equation*}
    And let $f_{n+1}(y) \defined f_n(y)$ for all $y \in Y_n$.

    To check that $f_{n+1}$ is an \kl{embedding}, let us first notice that
    words $f_{n+1}(x_i)$ and $f_{n+1}(x_j)$ are incomparable for the \kl{infix
    relation} when $i \neq j$, because the only factor of the form $c^{n+1}
    a^{n+1} b^t a^{n+1} c^{n+1}$ appears at the beginning of these words, and
    precisely encode $i$ (resp. $j$) in the value of $t$. Now, let us consider
    $y \in Y_n$ and $x_i \in X_{n+1}$ such that $y \preceq x_i$. Because $y \in
    D_i$, we have $f_n(y) \infleq f_{n+1}(x_i)$ by construction, and since
    $f_n(y) = f_{n+1}(y)$, we conclude that $f_{n+1}(y) \infleq f_{n+1}(x_i)$. 
    Conversely, assume that
    $f_{n+1}(y) \infleq f_{n+1}(x_i)$. Again, this means
    that $f_n(y) \infleq f_{n+1}(x_i)$. Now, the encoding
    is robust enough so that we can conclude
    $f_{n}(y) \infleq f_{n}(z)$ for some $z \in D_i$.
    By induction hypothesis, $y \preceq z$, hence, $y \preceq x_i$.

@@

As a consequence of \cref{infix-embedding:thm}, we cannot replay proofs of
\cref{prefixes:sec}, and will actually need to leverage some
regularity of the languages to obtain a characterization of well-quasi-ordered
languages under the infix relation. Let us first play this game for languages
that are recognized by finite automata.

\textbf{TODO: introduce finite automata and refer to the book.}.

\begin{theorem}
    \label{infix-finite-automata:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by a finite automaton.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

In order to prove \cref{infix-finite-automata:thm}, we will perform
some preliminary analysis on the structure of an automaton recognizing a
well-quasi-ordered language under the infix relation.

\begin{lemma}
    Let $A = (Q, \Sigma, \delta, q_0, F)$ be a finite automaton recognizing a
    language $L \subseteq \Sigma^*$. Assume furthermore that
    $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, for all words $u,w,v \in \Sigma^*$, and states
    $q, q_f \in Q \times F$ such that
    $q_0 \xrightarrow{u} q \xrightarrow{w} q \xrightarrow{v} q_f$,
    one of the following holds:
    \begin{itemize}
        \item There exists $k \in \Nat$
            such that $u$ is a \kl{suffix} of $w^k$,
        \item There exists $k \in \Nat$
            such that $v$ is a \kl{prefix} of $w^k$.
    \end{itemize}
\end{lemma}

\begin{lemma}
    Let $A = (Q, \Sigma, \delta, q_0, F)$ be a finite automaton recognizing a
    language $L \subseteq \Sigma^*$. Assume furthermore that
    $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, for all words $w_1, w_2 \in \Sigma^*$, and states
    $q, q_f \in Q \times F$ such that
    $q_0 \rightarrow^* q \xrightarrow{w} q \rightarrow^* q_f$
    and 
    $q_0 \rightarrow^* q \xrightarrow{w_2} q \rightarrow^* q_f$,
    there exists a word $w \in \Sigma^*$
    such that $w_1$ and $w_2$ are powers of $w$.
\end{lemma}

\begin{lemma}
    Let $A = (Q, \Sigma, \delta, q_0, F)$ be a finite automaton recognizing a
    language $L \subseteq \Sigma^*$. Assume furthermore that
    $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, for all words $u, w_1, w_2 \in \Sigma^*$, and states
    $q_1, q_2, q_f \in Q^2 \times F$,
    such that
    $q_0 \rightarrow^* q_1 \xrightarrow{w_1} q_1 \xrightarrow{u} q_2 \xrightarrow{w_2} q_2
    \rightarrow^* q_f$,
    there exists a $w \in \Sigma^*$ such that $w_1$, $w_2$, and $u$ are powers of $w$.
\end{lemma}

We are now ready to prove the main result.

\begin{proofof}{infix-finite-automata:thm}
    Let $w \in L$. Consider a run of the automaton on $w$. If $w$ is too long,
    then there exists a loop, and we can prove that $w = u x^p v$ for some
    short period $x$. By repeating the process for $u$ and $v$, we conclude
    that $w$ has the form $w = u_0 x_0^{p_0} u_1 x_1^{p_1} \cdots u_k x_k^{p_k}
    v_k$. Now, applying the loop merging lemma, we conclude that $x_0 = x_1 =
    \cdots = x_k$. Hence, $w = u_0 x^{p} v_k$. Finally, we know that either
    $u_0$ is a prefix of $x^{p}$ or that $v_k$ is a suffix of $x^{p}$. We
    conclude in particular that $\dwset[\infleq]{ \setof{ u x^p v }{ p \in \Nat
    }}$ is a finite union of \kl{chains}. Because $u,v,x$ are of bounded
    length, we conclude that $L$ itself included in a finite union of
    \kl{chains}, hence is a finite union of \kl{chains}.
\end{proofof}

\begin{corollary}
    Decidability?
\end{corollary}

@@

This section will use so-called \emph{amalgamation systems}.

>>>>

    Conversely, let us prove that \cref{infix-embedding-count:item} implies
    \cref{infix-embedding-embeds:item}. To that end, let us first define
    inductively a sequence $\seqof{X_n}$ of subsets of $X$ as follows: $X_0$ is
    the set of minimal elements of $X$, and $X_{n+1}$ is the set of minimal
    elements of $X \setminus \bigcup_{i \leq n} X_i$. Because initial segments
    of $X$ are finite, it is clear that $X = \bigcup_{n \in \Nat} X_n$. To
    simplify notations, we will also use $Y_n \defined \bigcup_{i \leq n} X_i$.
    We also let $L_0 \defined a b^* a$, and 
    $L_{n+1} \defined L_n \cup c^{n+1} b^* c^{n+1} (L_n c^{n+1})^*$.
    Now, let us inductively define \kl{embeddings} $f_n \colon Y_n \to
    L_n$.

    For the base case, let $\seqof{x_i}[i \in \Nat]$ be an enumeration of
    $X_0$, which is possible because $X$ is countable. We define $f_0 \colon
    X_0 \to \Sigma^*$ as follows: $f_0(x_i) \defined a b^i a$. It is immediate
    that $f_0$ is an \kl{embedding}. For the inductive step, let $\seqof{x_i}[i
    \in \Nat]$ be an enumeration of $X_{n+1}$, and assume that $f_n$ is an
    \kl{embedding} from $Y_n$ to $\Sigma^*$.
    Let us define $D_i \defined \setof{ y \in Y_n }{ y \preceq x_i }$
    for $i \in \Nat$, all of which are finite by assumption.
    We then let for all $i \in \Nat$:
    \begin{equation*}
        f_{n+1}(x_i) \defined 
        \underbrace{c^{n+1} b^i c^{n+1} \prod_{y \in D_i} \overbrace{f_n(y)}^{\in L_n} c^{n+1}}_{\in L_{n+1}}
        \quad .
    \end{equation*}
    And let $f_{n+1}(y) \defined f_n(y)$ for all $y \in Y_n$.

    To check that $f_{n+1}$ is an \kl{embedding}, let us first notice that
    words $f_{n+1}(x_i)$ and $f_{n+1}(x_j)$ are incomparable for the \kl{infix
    relation} when $i \neq j$, because the only factor of the form $c^{n+1}
    a^{n+1} b^t a^{n+1} c^{n+1}$ appears at the beginning of these words, and
    precisely encode $i$ (resp. $j$) in the value of $t$. Now, let us consider
    $y \in Y_n$ and $x_i \in X_{n+1}$ such that $y \preceq x_i$. Because $y \in
    D_i$, we have $f_n(y) \infleq f_{n+1}(x_i)$ by construction, and since
    $f_n(y) = f_{n+1}(y)$, we conclude that $f_{n+1}(y) \infleq f_{n+1}(x_i)$.
    Conversely, assume that $f_{n+1}(y) \infleq f_{n+1}(x_i)$. Again, this
    means that $f_n(y) \infleq f_{n+1}(x_i)$. Now, the encoding is robust
    enough so that we can conclude $f_{n}(y) \infleq f_{n}(z)$ for some $z \in
    D_i$ By induction hypothesis, $y \preceq z$, hence, $y \preceq x_i$.

@@

As a consequence of \cref{infix-embedding:thm}, we cannot replay proofs of
\cref{prefixes:sec}, and will actually need to leverage some regularity of the
languages to obtain a characterization of \kl{well-quasi-ordered} languages
under the \kl{infix relation}. Let us first play this game for languages that
are recognized by finite automata. We assume that the reader is familiar with
the notions of deterministic finite automaton and regular languages, and refer
to the book of Thomas for a comprehensive introduction to the subject
\cite{THOM97}. The main goal of the remainder of this section is to prove the
following \cref{infix-finite-automata:thm}.

\begin{theorem}[restate=infix-finite-automata:thm,label=infix-finite-automata:thm]
    Let $L \subseteq \Sigma^*$ be a language recognized by a finite automaton.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

\AP In order to prove \cref{infix-finite-automata:thm}, we will perform some
preliminary analysis on the structure of an automaton recognizing a
well-quasi-ordered language under the infix relation, which will be powered by
folklore results on \emph{periodic} words. Let us recall that a non-empty word
$w \in \Sigma^+$ is \intro(word){periodic} with period $x \in \Sigma^*$ if
there exists a $p \in \Nat$ such that $w \infleq x^p$. The \intro{periodic
length} of a word $u$ is the minimal length of a word $x$ such that $u$ is an
\kl{infix} of $x^p$ for some $p \in \Nat$ and $x \in \Sigma^+$. We will
essentially rely on the following result on periodic words.

\begin{lemma}
    \label{periodic-infixes:lem}
    Let $u,v \in \Sigma^*$ be two (non-empty) \kl{periodic words}
    having \kl{periodic lengths} $p$ and $q$ respectively.
    Then, if $u \infleq v$ and $\card{u} \geq \factorial[p]{p \times q}$,
    then $u$ and $v$ share the same \kl{periodic length}
    $p = q$.
\end{lemma}
\begin{proof}
    The fact that $u$ and $v$ are \kl{periodic length}
    respectively $p$ and $q$ translates into the fact that $u_{i+p} = u_i$ and
    $v_{i+q} = v_i$ for all indices $i \in \Nat$ such that those letters are
    well-defined.

    Now, assume that $u$ is an \kl{infix} of $v$, this provides the existence
    of a $k \in \Nat$ such that $u = v_{k} \cdots v_{k + \card{u} - 1}$. In
    particular, $v_{k+i+p} = v_{k+i}$ for all $i \in \Nat$ such that $k+i+p < k
    + \card{u}$. Since we also have $v_{k+i+q} = v_{k+i}$ for all $1 \leq i
    \leq \card{v} - k - q$. We conclude that both $u$ and $v$ are of
    \kl{periodic length} the greatest common divisor of $p$ and $q$, and by
    minimality of $q$ this must be equal to $q$ and to $p$.
\end{proof}

The reason why \kl{periodic words} built using a given period $x \in \Sigma^+$
are interesting for the \kl{infix relation} is that they naturally create
\kl{chains}. Indeed, if $x \in \Sigma^+$ is a finite word, then $\setof{x^p}{p
\in \Nat}$ is a \kl{chain} for the \kl{infix relation}. Note that in general,
the downwards closure of a chain is \emph{not} a chain. However, for the chains
generated using periodic words, the downwards closure
$\dwset[\infleq]{\setof{x^p}{p \in \Nat}}$ is a \emph{finite union} of
\kl{chains}. Because this set will appear in bigger equations,
we introduce the shorter notation 
$\intro*\InfPeriodChain{x}$ for the set of 
\kl{infixes} of words of the form $x^p$, where $p$ ranges over $\Nat$.

\begin{lemma}
    \label{inf-period-chain:lem}
    Let $x \in \Sigma^+$ be a word, and
    Then $\InfPeriodChain{x}$ is a finite union of \kl{chains}
    for the \kl{infix}, \kl{prefix} and \kl{suffix} relations 
    simultaneously.
\end{lemma}
\begin{proof}
    Let $x \in \Sigma^+$ be a word, and let $P_x$ be the (finite) set 
    of all \kl{prefixes} of $x$, and $S_x$ be the (finite)
    set of all \kl{suffixes} of $x$.
    Assume that $w \in \InfPeriodChain{x}$, then $w = u x^p v$ for some
    $u \in S_x$, $v \in P_x$, and $p \in \Nat$.
    We have proven that
    \begin{equation*}
        \InfPeriodChain{x} \subseteq \bigcup_{u \in P_x} \bigcup_{v \in S_x} u x^* v
        \quad .
    \end{equation*}

    Let us now demonstrate that for all $(u,v) \in S_x \times P_x$, the
    language $u x^* v$ is a \kl{chain} for the \kl{infix}, \kl{suffix} and \kl{prefix} relations.
    To that end,
    let $(u,v) \in S_x \times P_x$ and $\ell, k \in \Nat$ be such that $\ell <
    k$, let us prove that $u x^\ell v \infleq u x^k  v$. Because $v \prefleq
    x$, we know that there exists $w$ such that $vw = x$. In particular,
    $ux^\ell vw = u x^{\ell + 1}$, and because $\ell < k$, we conclude that $u
    x^{\ell + 1} \prefleq u x^k v$. By transitivity, $u x^\ell v \prefleq u x^k
    v$, and \emph{a fortiori}, $u x^\ell v \infleq u x^k v$. 
    Similarly, because $u \suffleq x$,  there exists $w$ such that $wu  = x$, 
    and we conclude that $u x^{\ell} v \suffleq w u x^\ell v = x^{\ell + 1} v \suffleq u x^k v$.
    \qedhere
\end{proof}

The following combinatorial lemma connects the property of being
\kl{well-quasi-ordered} to a property of the \kl{periodic lengths} of words in
a language, based on the assumption that some factors can be iterated.

\begin{lemma}
    \label{pumping-periods:lem}
    Let $L \subseteq \Sigma^*$ be a language
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Let $k \in \Nat$, $u_1, \cdots, u_{k+1} \in \Sigma^*$,
    and $v_1, \cdots, v_{k} \in \Sigma^+$
    be such that
    $w[\vec{X}] \defined (\prod_{i = 1}^k u_k v_k^{X_k}) u_{k+1}$
    belongs to $L$
    for arbitrarily large values of $\vec{X} \in \Nat^k$.
    Then, 
    there exists $x,y \in \Sigma^+$ of size 
    at most $\max \setof{\card{v_i}}{1 \leq i \leq k}$
    such that 
    one of the following holds for all
    $\vec{X} \in \Nat^{k}$:
    \begin{enumerate}
        \item $w[\vec{X}] \in u_1 \InfPeriodChain{x}$,
        \item $w[\vec{X}] \in \InfPeriodChain{x} u_{k+1}$,
        \item $w[\vec{X}] \in \InfPeriodChain{x} u_i \InfPeriodChain{y}$
            for some $1 \leq i \leq k + 1$.

    \end{enumerate}
\end{lemma}
\begin{proof}
    Let us construct a sequence of words $\seqof{w_i}[i \in \Nat]$, where $w_i
    = w[n_i, m_i]$ for some well-chosen indices $n_i, m_i \in \Nat$. The idea
    being that 
    if $w[n_i, m_i]$ is an \kl{infix} of $w[n_{j}, m_{j}]$,
    then one of $u_1$ or $u_3$ must be left untouched by the embedding,
    and the overlapping parts must be long enough to apply 
    \cref{periodic-infixes:lem}.
    In order to achieve this,
    we define
    $n_0$ and $m_0$ to be numbers greater than $\factorial[p]{\card{v_1} \times
    \card{v_2}}$ and such that $w[n_0, m_0]$ belongs to $L$. 
    Then, we define 
    $n_{i+1}$ and $m_{i+1}$ to be the smallest numbers greater than 
    $2 \times \card{w[n_i,m_i]}$ such that $w[n_{i+1}, m_{i+1}]$ belongs to $L$.

    Because $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}, there
    exists $i < j$ such that $w[n_i, m_i]$ is an \kl{infix} of $w[n_j, m_j]$.
    \textbf{TODO:} assume that we have a large intersection of $v_1$ and $v_2$,
    then we are in the latter cases. Otherwise, we are in the first case.
\end{proof}

\begin{proofof}{infix-finite-automata:thm}[main]
    Let $w \in L$, because $Q$ is finite, there exists
    a factorization of $w$
    into words $(\prod_{i = 1}^k u_i v_i) u_{k+1}$
    such that 
    for all $1 \leq i \leq k+1$, $\card{u_i} \leq \card{Q}$,
    for all $1 \leq i \leq k$, $1 \leq \card{v_i} \leq \card{Q}$,
    and satisfying 
    that $w[\vec{X}] \defined 
    (\prod_{i = 1}^k u_i v_i^{X_i}) u_{k+1}$
    belongs to $L$ for all choices of values $\vec{X} \in \Nat^k$.

    Applying \cref{pumping-periods:lem}, we conclude that 
    there exists $x,y \in \Sigma^+$ of size at most $\card{Q}$
    such that 
    $w \in u_1 \InfPeriodChain{x} \cup \InfPeriodChain{y} u_{k+1}
    \cup \bigcup_{1 \leq i \leq k+1} \InfPeriodChain{x} u_i \InfPeriodChain{y}$.
    In particular, we conclude that
    \begin{equation}
        \label{infix-automata:eq}
        L
        \subseteq
        \bigcup_{x,y,u\in \Sigma^{\leq \card{Q}}}
        u \InfPeriodChain{x}
        \cup 
        \InfPeriodChain{x} u
        \cup
        \InfPeriodChain{x} u \InfPeriodChain{y}
        \quad .
    \end{equation}

    Now, thanks to \cref{inf-period-chain:lem}
    this happens to be a finite union of \kl{chains}
    for the \kl{infix relation}.
\end{proofof}

\begin{corollary}
    Given a regular language $L$, it is decidable whether
    $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}
\begin{proof}
    It suffices to decide whether the equation 
    \cref{infix-automata:eq} holds for the language $L$.
    This is an inclusion of regular languages, which is decidable.
\end{proof}

@@

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument from \cref{pumping-periods:lem}, which made the proof
almost trivial. It turns out that there is a rather large family of systems for
which pumping arguments based on so-called \emph{minimal runs} exist: they are
called \emph{amalgamation systems}. Having done the heavy lifting on finite
automata, the rest of this section is mostly devoted to the introduction of
amalgamation systems and collecting the necessary pumping argument they enjoy.
Using this meta-proof, we will generalize \cref{infix-finite-automata:thm} to
context-free grammars, languages recognized by VASS, and more.
